Multi-reference pangenome graphs

### 2021-12-21 manuscript/main.tex
motivate

    \title[seqwish: inducing pangenome graphs]{seqwish: inducing pangenome graphs}
    \author[Garrison \textit{et~al}.]{
        Erik~Garrison\,$^{\text{\sfb 1}*}$,
        Andrea~Guarracino\,$^{\text{\sfb 2}}$
    }

@@

    \abstract{
        \textbf{Motivation:}
        Pangenome graphs provide a complete representation of the mutual alignment of collections of genomes.
        One of its most common forms is the variation graph model.
        A set of pairwise alignments imply a variation graph, but, to our knowledge, there are no methods to generate such a graph or lossless string graph from these alignments.\\
        \textbf{Results:}
        We wrote seqwish, a novel tool to induce a variation graph from a set of sequences and alignments between them.
        seqwish operates in external memory, by conversion of the alignment set into an alignment graph and the subsequent
        use of this graph in the elaboration of the variation graph including paths representing the input sequences.
        The resulting graph is a lossless representation of the input and its alignments.\\
        \textbf{Availability:}
        seqwish is published as free software under the MIT open source license.
        Source code and documentation are available at \url{https://github.com/ekg/seqwish}.
        seqwish can be installed via Bioconda \url{https://bioconda.github.io/recipes/seqwish/README.html} or GNU Guix \url{https://github.com/ekg/guix-genomics/blob/master/seqwish.scm}. \\
        \textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
%\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
    }

@@

    \section{Introduction}
    \label{sec:introduction}
    A pangenome models the full genomic information of a species or clase~\FIXME{PMID:32034321}.
    These models encode the mutual relationships between all the genomes represented, in contrast to reference-based approaches which relate sequences only to a particular genome, chosen as reference.
    A class of methods to represent pangenomes involves the \texttt{sequence graphs} \FIXME{PMID:2488477}, where similar orthologous regions between genomes collapse into a single representative sequence.
    In \texttt{node-labeled} sequence graphs, nodes indicate DNA sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
    A \texttt{bidirected} sequence graph represents both strands of DNA.
    On this model, variation graphs add the concept of \texttt{path} to embed linear sequences (e.g., genomes or haplotypes) into the graph\citep{Garrison:2018}.
    Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.

    \section{Model}
    \label{sec:model}
    A pangenome graph is a sequence model that encodes the mutual alignment of many genomes~\citep{Garrison_2019_thesis,Eizenga_2020}.
    In the variation graph, $V = (N, E, P)$, nodes $N = n_1\ldots n_{|N|}$ contain sequences of DNA.
    Each node $n_i$ has an identifier $i$ and an implicit reverse complement $\bar{n_i}$, and a node strand $s$ corresponds to one such orientations.
    Edges $E = e_1\ldots e_{|E|}$ represent ordered pairs of node strands: $e_i = ( s_a, s_b )$.
    Paths $P = p_1\ldots p_{|P|}$ describe walks over node strands: $p_i = s_1 \ldots s_{|p_i|}$.

>>>>

\title[Multi-reference pangenome graphs]{Multi-reference pangenome graphs}
\author[Garrison \textit{et~al}.]{
Erik~Garrison\,$^{\text{\sfb 1}*}$,
Andrea~Guarracino\,$^{\text{\sfb 2}}$
}

@@

\abstract{
\textbf{Motivation:}
Pangenome graphs provide a complete representation of the mutual alignment of collections of genomes.
They are commonly modeled as variation graphs.
A set of pairwise alignments imply a variation graph, but, to our knowledge, there are no methods to generate such a graph or lossless string graph from these alignments. \\
\textbf{Results:}
We design the \textit{seqwish} algorithm, which builds variation graph from a set of sequences and alignments between them.
Our seqwish implementation operates in external memory, using disk-backed implicit interval trees and a lock-free parallel union find algorithm to transform the alignment set into an alignment graph, then a sequence graph, and finally a variation graph with paths representing the input sequences.
%        use of this graph in the elaboration of the variation graph including paths representing the input sequences.
The resulting graph is a lossless representation of the input and its alignments.\\
\textbf{Availability:}
seqwish is published as free software under the MIT open source license.
Source code and documentation are available at \url{https://github.com/ekg/seqwish}.
seqwish can be installed via Bioconda \url{https://bioconda.github.io/recipes/seqwish/README.html} or GNU Guix \url{https://github.com/ekg/guix-genomics/blob/master/seqwish.scm}. \\
\textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
%\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
}

@@

\section{Introduction}
\label{sec:introduction}
A pangenome models the full genomic information of a species or clade~\FIXME{PMID:32034321}.
These models encode the mutual relationships between all the genomes represented, in contrast to reference-based approaches which relate sequences only to a particular ``reference'' genome.
%In principle, pangenome graphs can allow for a great reduction in reference bias, allowing any part of a pangenome to be considered in biological studies. %, and may provide a simpler representation of structurally-variable regions.
In a pangenome graph, variation can be understood in the context of any part of any included genome, which lets us avoid the problem of reference bias, or a limitation of analyses to sequences that are very similar to a chosen reference genome.

Surprisingly, existing methods to build pangenome graphs from large genome assemblies are reference-based.
Minigraph progressively builds a pangenome graph using by layering an ordered series of input genomes onto a specific reference.
Cactus depends on a single guide tree to structure pairwise alignments between genomes.
Both kinds of methods limit computational costs by reducing the number of pairwise comparisons, but in turn their results exhibit order and reference-dependency.

Although a reference-guided approach is pragmatic when we only have a single reference genome, in humans and other species recent advances in sequencing and assembly are driving the generation of many reference-quality assemblies.
Motivated by the need to generate ``multi-reference'' pangenome graphs, we present \textit{seqwish}, an algorithm that allows us to build pangenome graphs directly from a collection of sequences and pairwise alignments between them.
Here, we formally describe the algorithm, evaluate its basic combinatoric bounds and costs, and present applications of the method to pangenome building problems in a variety of species.
Our approach is both highly generic---in that we can apply it to truly any collection of sequences and alignments---but it easily scales to some of the largest species pangenome construction problems possible using publicly-available, high-quality genome assemblies.
%Not only is our approach highly generic,but it also efficiently scales to both large and complex pangenomes.
%with \textit{any} collection of alignments and sequences,

%Given recent advances in genome sequencing and assembly, we expect that soon, many species will have multiple reference-quality genomes.
%To build a pangenome model from such genomes that is truly unbiased, we need an approach to build a multi-reference 
%Motivated by the current availability of two reference-quality human genome assemblies 
%The inability to build pangenome graphs from

\section{Model}
    \label{sec:model}

A class of methods to represent pangenomes involves the \textit{sequence graphs} \FIXME{PMID:2488477}, where similar orthologous regions between genomes collapse into a single representative component in the graph.
In \textit{node-labeled} sequence graphs, nodes indicate DNA sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
A \textit{bidirected} sequence graph represents both strands of DNA.
On this model, variation graphs add the concept of \textit{path} to embed linear sequences (e.g., genomes or haplotypes) into the graph\citep{Garrison:2018}.
Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.

    A pangenome graph is a sequence model that encodes the mutual alignment of many genomes~\citep{Garrison_2019_thesis,Eizenga_2020}.
    In the variation graph, $V = (N, E, P)$, nodes $N = n_1\ldots n_{|N|}$ contain sequences of DNA.
    Each node $n_i$ has an identifier $i$ and an implicit reverse complement $\bar{n_i}$, and a node strand $s$ corresponds to one such orientations.
    Edges $E = e_1\ldots e_{|E|}$ represent ordered pairs of node strands: $e_i = ( s_a, s_b )$.
    Paths $P = p_1\ldots p_{|P|}$ describe walks over node strands: $p_i = s_1 \ldots s_{|p_i|}$.
